---
title: "California fire project"
output: html_document
date: "2025-06-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Inport the libraries and dataset
```{r,message=FALSE}
#install.packages("sf")
#install.packages("ggplot2")
#install.packages("tidyverse")
#install.packages("plotly")
setwd("~/Desktop/USRA/california fire")
library(sf)
library(ggplot2)
library(plotly)
library(raster)
library(tidyverse)
library(viridis)
library(gstat)
library(terra)

fire_data<-st_read("California_Historic_Fire_Perimeters.geojson")
fire_data<-
  fire_data %>% 
  filter(YEAR_>=1950) %>% 
  mutate(ALARM_DATE = substr(ALARM_DATE,5,16)) %>% 
  mutate(CONT_DATE = substr(CONT_DATE, 5, 16))

#introduce month to dataset

fire_data<-fire_data %>% 
  mutate(month = substr(ALARM_DATE,5,7))

fire_data<-fire_data %>% 
  mutate(season = case_when(
    month %in% c("Dec", "Jan", "Feb")  ~ "Winter",
    month %in% c("Mar", "Apr", "May")   ~ "Spring",
    month %in% c("Jun", "Jul", "Aug")   ~ "Summer",
    month %in% c("Sep", "Oct", "Nov") ~ "Fall",
    TRUE ~ NA_character_
  ))

fire_data$ALARM_DATE <- as.Date(fire_data$ALARM_DATE,
                                format = "%d %b %Y")
fire_data$CONT_DATE <- as.Date(fire_data$CONT_DATE,
                               format = "%d %b %Y")
fire_data<-fire_data %>% 
  mutate(time_diff = CONT_DATE-ALARM_DATE)




# full path to folder containing the files
shapefile_path <- "/Users/nancy/Desktop/USRA/california fire/tl_2023_us_state/tl_2023_us_state.shp"

# read california border map
stateshapes <- st_read(shapefile_path)
CA_shape<-stateshapes %>% filter(NAME=="California")
```

Seperate the data into four parts depending on their decades.
```{r,message=FALSE}
fires_1990s<-fire_data %>% filter(DECADES=="1990-1999")
fires_2000s<-fire_data %>% filter(DECADES=="2000-2009")
fires_2010s<-fire_data %>% filter(DECADES=="2010-2019")
fires_2020s <- fire_data %>% filter(DECADES == "2020-January 2025")

```


# Get the fire map of California of each deacades.  
Maps in 1990-1999.  
```{r}
ggplot()+
  geom_sf(data = fires_1990s,fill = "darkorange", color = "darkred")+
  geom_sf(data = CA_shape, fill = NA, color = "black", size = 1)+
  ggtitle("Fire Extents: 1990s") +
  theme_minimal()
```

Map in 2000-2009.  
```{r}
ggplot()+
  geom_sf(data = fires_2000s,fill = "darkorange", color = "darkred")+
  geom_sf(data = CA_shape, fill = NA, color = "black", size = 1)+
  ggtitle("Fire Extents: 2000s") +
  theme_minimal()
```

Map in 2010-2019.  
```{r}
ggplot()+
  geom_sf(data = fires_2010s,fill = "darkorange", color = "darkred")+
  geom_sf(data = CA_shape, fill = NA, color = "black", size = 1)+
  ggtitle("Fire Extents: 2010s") +
  theme_minimal()
```


Map of California fire in 2020s.
```{r}
#Map of California fire in 2020s 
ggplot()+
  geom_sf(data = fires_2020s,fill = "darkorange", color = "darkred")+
  geom_sf(data = CA_shape, fill = NA, color = "black", size = 1)+
  ggtitle("Fire Extents: 2020s") +
  theme_minimal()
```
  


# Fire area summary according to the decades.  


```{r}
area_summary_by_decades<-fire_data %>%
  st_drop_geometry()%>% 
  group_by(DECADES) %>% 
  summarise(average=mean(GIS_ACRES)) %>% 
  na.omit()
ggplot(data = area_summary_by_decades,aes(x = DECADES, y = average, group = 1))+
  geom_bar(stat = "identity",fill = "lightgrey")+
  geom_point(color = "blue")+
  geom_line(color = "red")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))+
  labs(title = "Decade average fire area")
```
# Fire area summary according to years.  
```{r}
area_summary_by_year<-fire_data %>% 
  st_drop_geometry()%>% 
  group_by(YEAR_) %>% 
  summarise(average=mean(GIS_ACRES)) %>% 
  na.omit()

ggplot(data = area_summary_by_year,aes(x = YEAR_, y = average))+
  geom_line(color = "blue")+
  geom_point(color = "red")+
  labs(title = "Averaged fire area by year")
```
#Analyze the length of the fire depending on the decades
```{r}
fire_timediff_by_decades<-fire_data %>% 
  st_drop_geometry() %>% 
  group_by(DECADES) %>% 
  summarise(avg_time_diff = mean(time_diff,na.rm = TRUE))

ggplot(fire_timediff_by_decades)+
  geom_line(aes(x = DECADES,y = avg_time_diff, group = 1))+
  labs(title = "Average time length of fire by year")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
  


fire_timediff_by_year<-fire_data %>% 
  st_drop_geometry() %>% 
  group_by(YEAR_) %>% 
  summarise(avg_time_diff = mean(time_diff,na.rm = TRUE))

ggplot(fire_timediff_by_year)+
  geom_line(aes(x = YEAR_,y = avg_time_diff, group = 1))+
  labs(title = "Average time length of fire by year")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
  


```



# Analyze the fire by seasons
```{r}


data_summary_by_decade_season<-fire_data %>% 
  st_drop_geometry() %>% 
  group_by(DECADES,season) %>% 
  summarise(count = n(),
            average_area = mean(GIS_ACRES)) %>% 
  arrange(DECADES,count,average_area) %>% 
  na.omit()
head(data_summary_by_decade_season, n =5)
```
## Plot of amount of the fire of each decades according to seasons
```{r}
ggplot(data = data_summary_by_decade_season, aes(group = season, color = season)) +
  geom_line(aes(x = DECADES, y = count, fill = season))+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))+
  labs(title = "Decades fire amount based on seasons")

ggplot(data = data_summary_by_decade_season, aes(group = season, color = season)) +
  geom_bar(aes(x = DECADES, y = count, fill = season), 
           stat = "identity", position = "stack")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))+
  labs(title = "Decades fire amount based on seasons")

```

## Plot of averaged area of each decades according to seasons
```{r}
ggplot(data = data_summary_by_decade_season,aes(group = season, color = season))+
  geom_line(aes(x = DECADES, y = average_area))+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))+
  labs(title = "Decades fire averaged area based on seasons")


ggplot(data = data_summary_by_decade_season) +
  geom_bar(aes(x = DECADES, y = average_area, fill = season), 
           stat = "identity", position = "stack")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))+
  labs(title = "Decades fire average area based on seasons")

```

# Summary of fire data according to seasons
```{r}
data_summary_by_season<-fire_data %>% 
  st_drop_geometry() %>% 
  group_by(season) %>% 
  summarise(count = n(),
            average_area = mean(GIS_ACRES,na.rm = TRUE)
            ) %>% 
  na.omit()

data_summary_by_season <- data_summary_by_season %>%
  mutate(season = factor(season, 
                         levels = c("Spring", "Summer", "Fall","Winter")))
```
  
Make a graph of the count and averaged  fire area according to seasons.    


```{r}
plot_ly() %>%
  add_bars(
    x = data_summary_by_season$season, 
    y = data_summary_by_season$count, 
    name = "Count",
    yaxis = "y1"
  ) %>%
  add_lines(
    x = data_summary_by_season$season, 
    y = data_summary_by_season$average_area, 
    name = "Average Area",
    yaxis = "y2"
  ) %>%
  layout(
    yaxis = list(title = "Count"),
    yaxis2 = list(title = "Average Area", overlaying = "y", side = "right"),
    xaxis = list(title = "Season"),
    title = "Counts & Average Area"
  )


```
# Analyze the causation of the fire
We can conclude that the lightening is the biggest natural reason for fire in CA.
```{r}
cause_summary<-fire_data %>% 
  st_drop_geometry() %>% 
  group_by(CAUSE) %>% 
  summarise(total_amount = sum(CAUSE)) %>% 
  arrange(-total_amount) %>% 
  na.omit()
head(cause_summary, n = 6)
```

```{r}
lightening_summarize<-fire_data %>% 
  st_drop_geometry() %>% 
  select(YEAR_,CAUSE) %>% 
  group_by(YEAR_,CAUSE) %>% 
  summarize(total = n())

YEAR_total <- lightening_summarize %>%
  group_by(YEAR_) %>%
  summarize(year_total = sum(total))


lightening_summarize <- lightening_summarize %>%
  left_join(YEAR_total, by = "YEAR_") %>% 
  mutate(proportion = total / year_total) %>% 
  filter(CAUSE == 1)

ggplot(data = lightening_summarize)+
  geom_line(aes(x = YEAR_, y = proportion*100, group= 1))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))+
  labs(title = "proportion of fire caused by lightening")


lightening_summarize_season<-fire_data %>% 
  st_drop_geometry() %>% 
  filter(CAUSE==1,!is.na(season)) %>% 
  select(DECADES,CAUSE,season) %>% 
  group_by(DECADES,season) %>% 
  summarise(season_total = n())

Decade_total_season<-lightening_summarize_season %>%
  group_by(DECADES) %>%
  summarize(decade_total = sum(season_total))
  
lightening_summarize_season <- lightening_summarize_season %>%
  left_join(Decade_total_season, by = "DECADES") %>%
  mutate(proportion = season_total / decade_total) %>% 
  filter(season=="Summer")



ggplot(data = lightening_summarize_season)+
  geom_line(aes(x = DECADES, y = proportion*100, group= 1))+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))+
  labs(title = "proportion of fire caused by summer lightening")

ggplot(data = lightening_summarize_season)+
  geom_line(aes(x = DECADES, y = season_total, group = 1))+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))+
  labs(title = "Amount of summer fire caused by lightening")

```
# Analyze the temperature change in CA from 1955 to 1985 to 2024.  
  
  
Analyze on 1955. 

```{r}
temperature_1955<-read.csv("temperature_195506-195508.csv")

avg_temp_1955<-temperature_1955 %>% 
  group_by(STATION) %>% 
  summarize(avg_temp = mean(TAVG),
            lat = mean(LATITUDE),
            lon = mean(LONGITUDE),
            ele = mean(ELEVATION)) %>% 
  na.omit()

avg_temp_sf_1955<-st_as_sf(avg_temp_1955,coords = c("lon","lat"), crs = 4326)

ggplot()+
  geom_sf(data = CA_shape)+
  geom_sf(data = avg_temp_sf_1955, 
          aes(color = avg_temp))+
  scale_color_gradient(low = "yellow", high = "red",
                        limits = c(35, 105))+
  labs(title = "Average temperature in 1955 July - 1955 Aug")

```
  
Analyze on 1985.   
```{r}
temperature_1985<-read.csv("temperature_198506-198508.csv")

avg_temp_1985<-temperature_1985 %>% 
  group_by(STATION) %>% 
  summarize(avg_temp = mean(TAVG),
            lat = mean(LATITUDE),
            lon = mean(LONGITUDE),
            ele = mean(ELEVATION)) %>% 
  na.omit()
avg_temp_sf_1985<-st_as_sf(avg_temp_1985,coords = c("lon","lat"), crs = 4326)
ggplot()+
  geom_sf(data = CA_shape)+
  geom_sf(data = avg_temp_sf_1985,aes(color = avg_temp))+
  scale_color_gradient(low = "yellow", high = "red",
                        limits = c(35, 105))+
  labs(title = "Average temperature in 1985 July - 1985 Aug")



```

Analyze on 2024. 
```{r}
temperature_2024<-read.csv("temperature_202406-202408.csv")

avg_temp_2024<-temperature_2024 %>% 
  group_by(STATION) %>% 
  summarize(avg_temp = mean(TAVG),
            lat = mean(LATITUDE),
            lon = mean(LONGITUDE),
            ele = mean(ELEVATION)) %>% 
  na.omit()
avg_temp_sf_2024<-st_as_sf(avg_temp_2024,coords = c("lon","lat"), crs = 4326)
ggplot()+
  geom_sf(data = CA_shape)+
  geom_sf(data = avg_temp_sf_2024,aes(color = avg_temp))+
  scale_color_gradient(low = "yellow", high = "red",
                        limits = c(35, 105))+
  labs(title = "Average temperature in 2024 May - 2024 Aug")

```
# Transfer the simple feature data frame into a raster to make the graph

```{r,message=FALSE}
fire_1955_lightening<-fire_data %>% 
  filter(YEAR_ == 1955) %>% 
  filter(CAUSE == 1) %>% 
  filter(CAUSE == 1, season == "Summer")

#transfer the sf dataframe into raster
temp_sp_1955 <- as_Spatial(avg_temp_sf_1955)

raster_1955 <- raster(extent(temp_sp_1955), res = 0.1)
crs(raster_1955) <- crs(temp_sp_1955)
# Convert empty raster to SpatialPixelsDataFrame
grid_1955 <- as(raster_1955, "SpatialPixelsDataFrame")

# Run IDW interpolation correctly
idw_result_1955 <- idw(formula = avg_temp ~ 1,
                  locations = temp_sp_1955,
                  newdata = grid_1955)

idw_raster_1955 <- raster(idw_result_1955)

idw_df_1955 <- as.data.frame(idw_raster_1955, xy = TRUE)


```

```{r,message=FALSE}
fire_1985_lightening<-fire_data %>% 
  filter(YEAR_ == 1985) %>% 
  filter(CAUSE == 1) %>% 
  filter(CAUSE == 1, season == "Summer")


#transfer the sf dataframe into raster
temp_sp_1985 <- as_Spatial(avg_temp_sf_1985)

raster_1985 <- raster(extent(temp_sp_1985), res = 0.1)
crs(raster_1985) <- crs(temp_sp_1955)
# Convert empty raster to SpatialPixelsDataFrame
grid_1985 <- as(raster_1985, "SpatialPixelsDataFrame")

# Run IDW interpolation correctly
idw_result_1985 <- idw(formula = avg_temp ~ 1,
                  locations = temp_sp_1985,
                  newdata = grid_1985)

idw_raster_1985 <- raster(idw_result_1985)


idw_df_1985 <- as.data.frame(idw_raster_1985, xy = TRUE)


```

```{r,message=FALSE}
fire_2024_lightening<-fire_data %>% 
  filter(YEAR_ == 2024) %>% 
  filter(CAUSE == 1, season == "Summer")

temp_sp_2024 <- as_Spatial(avg_temp_sf_2024)
raster_2024 <- raster(extent(temp_sp_2024), res = 0.1)
crs(raster_2024) <- crs(temp_sp_2024)
# Convert empty raster to SpatialPixelsDataFrame
grid_2024 <- as(raster_2024, "SpatialPixelsDataFrame")

# Run IDW interpolation correctly
idw_result <- idw(formula = avg_temp ~ 1,
                  locations = temp_sp_2024,
                  newdata = grid_2024)

idw_raster <- raster(idw_result)

idw_df <- as.data.frame(idw_raster, xy = TRUE)
```

## Making plot based on the temperature raster data.
```{r}
ggplot() +
  geom_raster(data = idw_df_1955, aes(x = x, y = y, fill = var1.pred)) +
  geom_sf(data = CA_shape, fill = NA, color = "black", size = 0.5) +
  geom_sf(data = fire_1955_lightening, fill = NA, color = "black")+
  coord_sf() +
  theme_minimal() +
  labs(title = "Avg Temp & Fire in 1955")+
  scale_fill_gradient(name = "Temp (F)", low = "yellow", high = "red",limits = c(35, 105))


ggplot() +
  geom_raster(data = idw_df_1985, aes(x = x, y = y, fill = var1.pred)) +
  geom_sf(data = CA_shape, fill = NA, color = "black", size = 0.5) +
  geom_sf(data = fire_1985_lightening, fill = NA, color = "black")+
  coord_sf() +
  theme_minimal() +
  labs(title = "Avg Temp & Fire in 1985")+
  scale_fill_gradient(name = "Temp (F)", low = "yellow", high = "red",limits = c(35, 105))


ggplot() +
  geom_raster(data = idw_df, aes(x = x, y = y, fill = var1.pred)) +
  geom_sf(data = CA_shape, fill = NA, color = "black", size = 0.5) +
  geom_sf(data = fire_2024_lightening, fill = NA, color = "black")+
  coord_sf() +
  theme_minimal() +
  labs(title = "Avg Temp & Fire in 2024")+
  scale_fill_gradient(name = "Temp (F)", low = "yellow", high = "red",limits = c(35, 105))


```

```{r}
temp_rcp585<-rast(
  "tas_Amon_GFDL-ESM4_ssp585_r1i1p1f1_gr1_201501-210012.nc"
  )
temp_rcp585<- shift(temp_rcp585, dx = -180)

california<-vect(CA_shape)

california <- project(california, crs(temp_rcp585))
extent_california <- ext(california)

cropped_raster <- crop(temp_rcp585,extent_california )

california_raster <- mask(cropped_raster, california)

land_only_raster <- mask(california_raster, california)

raster_date<-(time(california_raster))

raster_date<-as.character(raster_date)

california_df <- as.data.frame(california_raster, xy = TRUE, na.rm = TRUE)

names(california_df) <- c("x", "y",raster_date)  # Rename to match aes(x, y, fill)


california_df[3:1034] <- (california_df[3:1034] - 273.15) * 9/5 + 32

california_df_longer<-california_df %>% 
  pivot_longer(cols = raster_date,
               names_to = "Date",
               values_to = "Temperature")

temp_2024_summer<-california_df_longer %>% 
  filter(Date=="2024-06-16"|Date == "2024-07-16"|Date=="2024-08-16") %>% 
  group_by(x,y) %>% 
  summarise(avg_temp = mean(Temperature))

fire_2024_jun<-fire_data %>% 
  filter(YEAR_==2024,season == "Summer",CAUSE == 1)

ggplot() +
  geom_raster(data = temp_2024_summer, aes(x = x, y =y, fill = avg_temp)) +
  geom_sf(data = CA_shape,fill = NA, color = "black")+
  geom_sf(data = fire_2024_jun)+
  coord_sf() + 
  scale_fill_gradient(name = "Temp (F)", low = "yellow", high = "red")+
  labs(
    title = "Temperature & Fire in California in 2024 summer",
    x = "Longitude", y = "Latitude"
  )+
  theme_minimal(base_size = 14)


plot(temp_rcp585[[1]], main = "Temperature in 2024 summer")
plot(CA_shape$geometry,add = TRUE)

library(dplyr)

california_df %>%
  mutate(region = ifelse(y > 37, "North", "South")) %>%
  group_by(region) %>%
  summarise(mean_temp = mean(`2024-06-16`, na.rm = TRUE))

library(maps)
# conus_maps# conus_sf
conus_sf <- st_as_sf(map("state", plot = FALSE, fill = TRUE))
conus_otl_sf <- st_geometry(conus_sf)
US<-world[world$name_long == "United States",]
plot(US$geom)
plot(conus_otl_sf,add = TRUE)



```

