---
title: "california fire"
output: html_document
date: "2025-08-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import California's shape
```{r}
shapefile_path <- "/Users/nancy/Desktop/USRA/california fire/tl_2023_us_state/tl_2023_us_state.shp"
stateshapes <- st_read(shapefile_path)
CA_shape<-stateshapes %>% filter(NAME=="California")
CA_vect<-vect(CA_shape)
```


getting all the libraries
```{r}
library(tidycensus)
library(dplyr)
library(sf)
library(rvest)
library(dplyr)
library(stringr)
library(sf)
library(ggplot2)

library(sf)
library(terra)
library(tidyverse)
library(jsonlite)
library(sp)
library(gstat)
```

getting the data for population data
```{r}
census_api_key("fd0ec10fccaf6f825d7e50d42fc14e0d06d2bf14", install = TRUE)

# Pull total population (variable "B01003_001") from 2020 ACS
ca_population <- get_acs(
  geography = "tract",
  state = "CA",
  variables = "B01003_001",  # Total population
  year = 2023,
  geometry = FALSE
)

housing_data <- get_acs(
  geography = "tract",
  state = "CA",
  variables = "B25001_001",   # Total housing units
  year = 2023,
  geometry = FALSE
)

# Median housing value for CA tracts
value_data <- get_acs(
  geography = "tract",
  state = "CA",
  variables = "B25077_001",  # Median value of owner-occupied homes
  year = 2023,
  geometry = FALSE
)

library(tigris)
ca_tracts <- tracts(state = "CA", cb = TRUE, class = "sf")

CA_population_data <- ca_tracts %>% 
  left_join(ca_population, by = "GEOID") %>% 
  left_join(housing_data, by = "GEOID") %>% 
  left_join(value_data, by = "GEOID")

CA_population_data <- CA_population_data %>%
  group_by(NAMELSADCO) %>%  
  mutate(estimate = ifelse(is.na(estimate), mean(estimate, na.rm = TRUE), estimate)) %>%
  ungroup()

CA_population_data<-CA_population_data %>% 
  select(GEOID,estimate.x, estimate.y, estimate)
colnames(CA_population_data)[2:4]<-c("population","building_amount","building_value_median")
ca_pop_proj <- st_transform(CA_population_data, crs = 3310)
ca_pop_proj$area_m2 <- st_area(ca_pop_proj)
cell_size_m <- 0.05 * 111320  # 1 degree = 111.32 km
cell_area_m2 <- cell_size_m^2
ca_pop_proj$amount_grids<-ceiling(ca_pop_proj$area_m2/cell_area_m2)


st_write(CA_population_data,"~/Desktop/californiaUSRA/california population distribution.geojson")

```


Getting the historical climate data including the 
```{r}
token <- "IgBwexLSxntlocXcGGjgQKRNksaehRlM"  # your token

# Step 2: Download data

get_end_day <- function(month) {
  if (month == "02"){
    return ("28")
  }
  else if(month %in% c("06","04","09","11")) {
    return("30")
  } else 
    return("31")
}


years <- 2013:2024
months<-c(paste0(0,1:9),"10","11","12")
max_retries <- 3
limit <- 1000  
all_data <- list()
  
for (year in years) {
    for (month in months) {
      success <- FALSE
      offset <- 0
      collected_data <- list()
      
      repeat {
        for (attempt in 1:max_retries) {
          res <- tryCatch({
            GET(
              url = "https://www.ncei.noaa.gov/cdo-web/api/v2/data",
              query = list(
                datasetid = "GSOM",
                datatypeid = "TAVG",
                locationid = "FIPS:06",
                startdate = paste0(year, "-", month, "-01"),
                enddate = paste0(year, "-", month, "-", get_end_day(month)),
                units = "standard",
                limit = limit,
                offset = offset,
                includemetadata = TRUE
              ),
              add_headers(token = token)
            )
          }, error = function(e) NULL)
          
          if (!is.null(res)) {
            text <- content(res, "text", encoding = "UTF-8")
            
            if (!grepl("^\\s*<", text)) {
              content_json <- fromJSON(text)
              results <- content_json$results
              
              if (!is.null(results)) {
                collected_data <- append(collected_data, list(results))
                message(paste("✅ Year", year, "Month", month, 
                              ": page offset", offset, "collected"))
                offset <- offset + limit  # next page
                success <- TRUE
                break  # if succeed, then break
              } else {
                message(paste("⚠️ Year", year, "Month", month, 
                              ": no more data (offset", offset, ")"))
                success <- TRUE
                break  # if no data, break
              }
            } else {
              message(paste("⛔ Year", year, "Month", month, 
                            ": HTML returned (offset", offset, 
                            "Attempt", attempt, ")"))
            }
          } else {
            message(paste("❌ Year", year, "Month", month, 
                          ": Request failed (offset", offset, 
                          "Attempt", attempt, ")"))
          }
          
          Sys.sleep(runif(1, 3, 6))  # wait before retry
        }
        
        if (!success) {
          message(paste("🚫 Year", year, "Month", month, 
                        ": failed after", max_retries, 
                        "attempts (offset", offset, ")"))
          break  # fail for 3 times, break
        }
        
        if (length(results) < limit) break
      }
      
      # combine all the data
      if (length(collected_data) > 0) {
        all_data[[paste0(year, "-", month)]] <- do.call(rbind, collected_data)
      }
    }
  }

df <- bind_rows(all_data)

# get the information for precipitation-------------------------------------------------------------------------


max_retries <- 3
limit <- 1000  
all_data_prcp <- list()

for (year in years) {
  for (month in months) {
    success <- FALSE
    offset <- 0
    collected_data <- list()
    
    repeat {
      for (attempt in 1:max_retries) {
        res <- tryCatch({
          GET(
            url = "https://www.ncei.noaa.gov/cdo-web/api/v2/data",
            query = list(
              datasetid = "GSOM",
              datatypeid = "PRCP",
              locationid = "FIPS:06",
              startdate = paste0(year, "-", month, "-01"),
              enddate = paste0(year, "-", month, "-", get_end_day(month)),
              units = "standard",
              limit = limit,
              offset = offset,
              includemetadata = TRUE
            ),
            add_headers(token = token)
          )
        }, error = function(e) NULL)
        
        if (!is.null(res)) {
          text <- content(res, "text", encoding = "UTF-8")
          
          if (!grepl("^\\s*<", text)) {
            content_json <- fromJSON(text)
            results <- content_json$results
            
            if (!is.null(results)) {
              collected_data <- append(collected_data, list(results))
              message(paste("✅ Year", year, "Month", month, 
                            ": page offset", offset, "collected"))
              offset <- offset + limit  # next page
              success <- TRUE
              break  # if succeed, then break
            } else {
              message(paste("⚠️ Year", year, "Month", month, 
                            ": no more data (offset", offset, ")"))
              success <- TRUE
              break  # if no data, break
            }
          } else {
            message(paste("⛔ Year", year, "Month", month, 
                          ": HTML returned (offset", offset, 
                          "Attempt", attempt, ")"))
          }
        } else {
          message(paste("❌ Year", year, "Month", month, 
                        ": Request failed (offset", offset, 
                        "Attempt", attempt, ")"))
        }
        
        Sys.sleep(runif(1, 3, 6))  # wait before retry
      }
      
      if (!success) {
        message(paste("🚫 Year", year, "Month", month, 
                      ": failed after", max_retries, 
                      "attempts (offset", offset, ")"))
        break  # fail for 3 times, break
      }
      
      if (length(results) < limit) break
    }
    
    # combine all the data
    if (length(collected_data) > 0) {
      all_data_prcp[[paste0(year, "-", month)]] <- do.call(rbind, collected_data)
    }
  }
}
prcp_df<-bind_rows(all_data_prcp)



data_all<-df %>% 
  left_join(prcp_df,by = c("date","station"))
data_all$year <- substr(df$date, 1, 4)
data_all$month <- substr(df$date, 6, 7)
data_all<-data_all %>% 
  mutate(temp = value.x, prcp = value.y)


# Add information for stations and data cleaning-------------------------------------------------------------------------

station_ids <- unique(data_all$station)


offset <- 0
limit <- 1000
stations_all <- list()

repeat {
  res <- GET(
    url = "https://www.ncei.noaa.gov/cdo-web/api/v2/stations",
    query = list(
      locationid = "FIPS:06",   # California
      datasetid = "GSOM",       # Monthly summary dataset
      limit = limit,
      offset = offset
    ),
    add_headers(token = token)
  )
  
  content_text <- content(res, "text", encoding = "UTF-8")
  
  if (grepl("^\\s*<", content_text)) {
    cat("❌ HTML error at offset", offset, "\n")
    break
  }
  
  json_data <- fromJSON(content_text)
  
  if (is.null(json_data$results) || length(json_data$results) == 0) {
    cat("✅ No more stations found. Done.\n")
    break
  }
  
  stations_all[[length(stations_all) + 1]] <- json_data$results
  cat("📦 Retrieved", length(json_data$results), "stations at offset", offset, "\n")
  
  offset <- offset + limit
  Sys.sleep(0.5)  # avoid rate limiting
}


station_info<-rbind(stations_all[[1]],stations_all[[2]],stations_all[[3]]) %>% 
  as.data.frame()
station_df<-station_info %>% filter(id %in% station_ids)

temp_data<-data_all %>% left_join(station_info, by = c("station" = "id"))

data_sf<-temp_data %>% 
  select(year,month,elevation,temp,prcp,longitude,latitude)


write.csv(data_sf, "california_temp_n_prcp_with_location_2013-2024.csv", row.names = FALSE)

```

Get the climate data as raster format
```{r}
# elevation --------------------------------------
library(elevatr)
elev_raster <- get_elev_raster(locations = CA_shape, z = 8, clip = "bbox")
plot(elev_raster)

crs(elev_raster)<-crs(CA_vect)
elev_raster<-rast(elev_raster)
elev_raster<-crop(elev_raster,CA_vect)
elev_raster <- mask(elev_raster, CA_vect)
plot(elev_raster)

elev_raster<-resample(elev_raster,template_raster,method = "bilinear")
elev_df<-as.data.frame(elev_raster,xy = TRUE)
elev_df <- elev_df %>% 
  rename("elevation" = colnames(elev_df[3]))

# temperature --------------------------------------
temp_n_prcp<-read.csv("~/Desktop/USRA/california fire/california_temp_n_prcp_with_location_2013-2024.csv")
temperature_data<-temp_n_prcp %>% dplyr::select(-prcp)
#Rasterize it
temperature_sf<-st_as_sf(temperature_data,coords = c("latitude","longitude"),crs = 4326)

rasterization_kriging <- function(data) {
  # Keep longitude and latitude as columns before setting coordinates
  data$longitude_val <- data$longitude
  data$latitude_val <- data$latitude
  
  # Generate prediction grid
  grid_lon <- seq(min(data$longitude), max(data$longitude), by = 0.1)
  grid_lat <- seq(min(data$latitude), max(data$latitude), by = 0.1)
  grid_df <- expand.grid(lon = grid_lon, lat = grid_lat)
  
  # Convert to spatial
  coordinates(data) <- ~longitude + latitude
  coordinates(grid_df) <- ~lon + lat
  gridded(grid_df) <- TRUE
  
  # IDW interpolation of elevation to fill grid
  idw_ele <- idw(elevation ~ 1, locations = data, newdata = grid_df, idp = 2.0)
  idw_ele_df <- as.data.frame(idw_ele)
  names(idw_ele_df)[names(idw_ele_df) == "var1.pred"] <- "elevation"
  
  prediction_grid_df <- as.data.frame(grid_df)
  prediction_grid_df$elevation <- idw_ele_df$elevation
  
  # Add longitude and latitude to prediction grid as needed for regression
  prediction_grid_df$longitude_val <- coordinates(grid_df)[,1]
  prediction_grid_df$latitude_val <- coordinates(grid_df)[,2]
  
  coordinates(prediction_grid_df) <- ~lon + lat
  gridded(prediction_grid_df) <- TRUE
  
  # Fit variogram using elevation + lat/lon as predictors
  vgm_model <- variogram(temp ~ elevation + longitude_val + latitude_val, data = data)
  fitted_vgm <- fit.variogram(vgm_model, model = vgm("Sph", nugget = 0.1))
  
  # Kriging regression with consistent formula
  kriging_result <- krige(temp ~ elevation + longitude_val + latitude_val,
                          locations = data,
                          newdata = prediction_grid_df,
                          model = fitted_vgm)
  
  # Convert to raster and post-process
  kriging_rast <- rast(kriging_result)[[1]]
  crs(kriging_rast)<-crs(elev_raster)
  kriging_rast <- resample(kriging_rast, elev_raster, method = "bilinear")
  cropped_temp <- crop(kriging_rast, CA_vect)
  masked_temp <- mask(cropped_temp, CA_vect)
  
  return(masked_temp)
}

years<-2013:2024
mons<-1:12
temp_list <- list()

for (yr in years) {
  for (mon in mons) {
    temp <- temperature_data %>%
      filter(year == yr, month == mon) %>%
      distinct(longitude, latitude, .keep_all = TRUE)
    
    key <- paste0(as.character(yr), sprintf("%02d", mon))
    temp_list[[key]] <- rasterization_kriging(temp)
  }
}
flat_temp_list <- unlist(temp_list, recursive = TRUE)
temp_stack <- rast(flat_temp_list)

#temp_stack<-resample(temp_stack,template_raster,method = "bilinear")

temp_df <- as.data.frame(temp_stack, xy = TRUE)
temp_df <- temp_df %>%
  pivot_longer(
    cols = c(-x, -y),
    names_to = "time",
    values_to = "value"
  ) %>%
  mutate(
    year = as.numeric(substr(time, 1, 4)),
    month = as.numeric(substr(time, 5, 6))
  ) %>%
  rename(temperature = value)

# precipitation --------------------------------------

precipitation_data<-temp_n_prcp %>% dplyr::select(-temp)


rasterization_kriging_prcp <- function(data) {  # Keep longitude and latitude as columns before setting coordinates
  data$longitude_val <- data$longitude
  data$latitude_val <- data$latitude
  
  # Generate prediction grid
  grid_lon <- seq(min(data$longitude), max(data$longitude), by = 0.1)
  grid_lat <- seq(min(data$latitude), max(data$latitude), by = 0.1)
  grid_df <- expand.grid(lon = grid_lon, lat = grid_lat)
  
  # Convert to spatial
  coordinates(data) <- ~longitude + latitude
  coordinates(grid_df) <- ~lon + lat
  gridded(grid_df) <- TRUE
  
  # IDW interpolation of elevation to fill grid
  idw_ele <- idw(elevation ~ 1, locations = data, newdata = grid_df, idp = 2.0)
  idw_ele_df <- as.data.frame(idw_ele)
  names(idw_ele_df)[names(idw_ele_df) == "var1.pred"] <- "elevation"
  
  prediction_grid_df <- as.data.frame(grid_df)
  prediction_grid_df$elevation <- idw_ele_df$elevation
  
  # Add longitude and latitude to prediction grid as needed for regression
  prediction_grid_df$longitude_val <- coordinates(grid_df)[,1]
  prediction_grid_df$latitude_val <- coordinates(grid_df)[,2]
  
  coordinates(prediction_grid_df) <- ~lon + lat
  gridded(prediction_grid_df) <- TRUE
  
  # Fit variogram using elevation + lat/lon as predictors
  vgm_model <- variogram(prcp ~ elevation + longitude_val + latitude_val, data = data)
  fitted_vgm <- fit.variogram(vgm_model, model = vgm("Sph", nugget = 0.1))
  
  # Kriging regression with consistent formula
  kriging_result <- krige(prcp ~ elevation + longitude_val + latitude_val,
                          locations = data,
                          newdata = prediction_grid_df,
                          model = fitted_vgm)
  
  # Convert to raster and post-process
  kriging_rast <- rast(kriging_result)[[1]]
  crs(kriging_rast)<-crs(elev_raster)
  kriging_rast <- resample(kriging_rast, elev_raster, method = "bilinear")
  cropped_temp <- crop(kriging_rast, CA_vect)
  masked_temp <- mask(cropped_temp, CA_vect)
  
  return(masked_temp)
}

prcp_list <- list()

for (yr in years) {
  for (mon in mons) {
    temp <- precipitation_data %>%
      filter(year == yr, month == mon) %>%
      distinct(longitude, latitude, .keep_all = TRUE) %>% 
      na.omit()
    
    key <- paste0(as.character(yr), sprintf("%02d", mon))
    prcp_list[[key]] <- rasterization_kriging_prcp(temp)
  }
}
flat_prcp_list <- unlist(prcp_list, recursive = TRUE)
prcp_stack <- rast(flat_prcp_list)


prcp_df <- as.data.frame(prcp_stack, xy = TRUE)
prcp_df <- prcp_df %>%
  pivot_longer(
    cols = c(-x, -y),
    names_to = "time",
    values_to = "value"
  ) %>%
  mutate(
    year = as.numeric(substr(time, 1, 4)),
    month = as.numeric(substr(time, 5, 6))
  ) %>%
  rename(precipitation = value)

prcp_df <- prcp_df %>%
  mutate(precipitation = ifelse(precipitation < 0, 0, precipitation))

#there are some duplicated parts
#take the average of the same position and same time
prcp_df <- prcp_df %>%
  group_by(x, y, year, month) %>%
  summarise(precipitation = mean(precipitation, na.rm = TRUE), .groups = "drop")



temp_prcp_n_elev<-temp_df %>% 
  left_join(prcp_df, by = c("year", "month", "x", "y")) %>% 
  left_join(elev_df, by = c("x", "y")) %>% 
  na.omit()
temp_prcp_n_elev<-temp_prcp_n_elev %>% 
  mutate(precipitation =  ifelse(precipitation<0, 0, precipitation))
temp_prcp_n_elev<-temp_prcp_n_elev %>% 
  filter(elevation>=-86)
temp_prcp_n_elev<-temp_prcp_n_elev %>% dplyr::select(-time.x,-time.y)

write_csv(temp_prcp_n_elev, "/Users/nancy/Desktop/californiaUSRA/california_natural_factors.csv")


```



Estimate the losses in monetary unit
```{r}
loss_data<-read.csv("~/Desktop/USRA/california fire/loss/mapdataall.csv")

#data_manipulation of the data time

loss_data$incident_dateonly_created<-loss_data$incident_dateonly_created %>% 
  as.Date()
loss_data$incident_dateonly_extinguished<-as.Date(
  ifelse(loss_data$incident_dateonly_extinguished=="",
         substr(loss_data$incident_date_last_update,1,10),
         loss_data$incident_dateonly_extinguished),
  format = "%Y-%m-%d")
loss_data$duration<-loss_data$incident_dateonly_extinguished-
  loss_data$incident_dateonly_created
loss_data$start_year<-format(loss_data$incident_dateonly_created, "%Y") %>% as.numeric()
loss_data$start_mon<-format(loss_data$incident_dateonly_created, "%m") %>% as.numeric()

get_fire_resources_and_damage <- function(url) {
  page <- read_html(url)
  
  #Extract Burned Area
  area <- page %>%
    html_nodes(xpath = "//*[contains(text(), ' Acres')]") %>%  # any element containing "Acres"
    html_text(trim = TRUE) %>%
    .[1] %>%                                # take the first occurrence
    str_extract("[0-9,]+") %>%              # extract numeric part
    str_replace_all(",", "") %>%            # remove commas
    as.numeric()
  
  # Extract resources assigned
  personnel <- page %>% html_nodes(xpath = "//div[contains(text(), 'Personnel')]/preceding-sibling::div") %>% 
    html_text(trim = TRUE) %>% as.numeric()
  helicopters <- page %>% html_nodes(xpath = "//div[contains(text(), 'Helicopters')]/preceding-sibling::div") %>% 
    html_text(trim = TRUE) %>% as.numeric()
  engines <- page %>% html_nodes(xpath = "//div[contains(text(), 'Engines')]/preceding-sibling::div") %>% 
    html_text(trim = TRUE) %>% as.numeric()
  dozer <- page %>% html_nodes(xpath = "//div[contains(text(), 'Dozer')]/preceding-sibling::div") %>% 
    html_text(trim = TRUE) %>% as.numeric()
  Water_Tenders <- page %>% html_nodes(xpath = "//div[contains(text(), 'Water Tenders')]/preceding-sibling::div") %>% 
    html_text(trim = TRUE) %>% as.numeric()
  crews <- page %>% html_nodes(xpath = "//div[contains(text(), 'Crews')]/preceding-sibling::div") %>% 
    html_text(trim = TRUE) %>% as.numeric()
  Other_Assigned <- page %>% html_nodes(xpath = "//div[contains(text(), 'Other Assigned')]/preceding-sibling::div") %>% 
    html_text(trim = TRUE) %>% as.numeric()
  
  # Extract damage assessment
  structures_damaged <- page %>% html_nodes(xpath = "//div[contains(text(), 'Structures Damaged')]/preceding-sibling::div") %>% 
    html_text(trim = TRUE) %>% as.numeric()
  structures_destroyed <- page %>% 
    html_nodes(xpath = "//div[contains(text(), 'Structures Destroyed')]/preceding-sibling::div") %>% 
    html_text(trim = TRUE) %>% 
    gsub(",", "", .) %>%       # remove commas
    as.numeric()
  civilian_fatalities <- page %>% html_nodes(xpath = "//div[contains(text(), 'Confirmed Civilian Fatalities')]/preceding-sibling::div") %>% 
    html_text(trim = TRUE) %>% as.numeric()
  civilian_injuries <- page %>% html_nodes(xpath = "//div[contains(text(), 'Confirmed Civilian Injuries')]/preceding-sibling::div") %>% 
    html_text(trim = TRUE) %>% as.numeric()
  firefighter_fatalities<-page %>% html_nodes(xpath = "//div[contains(text(), 'Confirmed Firefighter Fatalities')]/preceding-sibling::div") %>% 
    html_text(trim = TRUE) %>% as.numeric()
  firefighter_injuries <- page %>% html_nodes(xpath = "//div[contains(text(), 'Confirmed Firefighter Injuries')]/preceding-sibling::div") %>% 
    html_text(trim = TRUE) %>% as.numeric()
  
  # Return as a data frame row
  return(data.frame(
    area_acres = ifelse(length(area) == 0, NA, area),
    
    personnel = ifelse(length(personnel) == 0, NA, personnel),
    helicopters = ifelse(length(helicopters) == 0, NA, helicopters),
    engines = ifelse(length(engines) == 0, NA, engines),
    dozer = ifelse(length(dozer) == 0, NA, dozer),
    Water_Tenders = ifelse(length(Water_Tenders) == 0, NA, Water_Tenders),
    crews = ifelse(length(crews) == 0, NA, crews),
    Other_Assigned = ifelse(length(Other_Assigned) == 0, NA, Other_Assigned),
    
    structures_damaged = ifelse(length(structures_damaged) == 0, NA, structures_damaged),
    structures_destroyed = ifelse(length(structures_destroyed) == 0, NA, structures_destroyed),
    civilian_fatalities = ifelse(length(civilian_fatalities) == 0, NA, civilian_fatalities),
    civilian_injuries = ifelse(length(civilian_injuries) == 0, NA, civilian_injuries),
    firefighter_fatalities = ifelse(length(firefighter_fatalities) == 0, NA, firefighter_fatalities),
    firefighter_injuries = ifelse(length(firefighter_injuries) == 0, NA, firefighter_injuries),
    stringsAsFactors = FALSE
  ))
}

#get the details for each fire according to their links
#it takes approximately 30 mins
loss_link <- loss_data$incident_url
get_fire_resources_and_damage("https://www.fire.ca.gov/incidents/2025/7/9/orleans-complex")
result <- lapply(loss_link, get_fire_resources_and_damage)

#combine the loss data with the fire data
loss_df<-bind_rows(result)
loss_df[is.na(loss_df)] <- 0
loss_data_combined<-bind_cols(loss_data,loss_df)
loss_data_combined$loss_estimated<-
  loss_data_combined$structures_damaged*0.4+
  loss_data_combined$structures_destroyed*0.8+
  loss_data_combined$civilian_fatalities*13.7+
  loss_data_combined$civilian_injuries*6.85+
  loss_data_combined$firefighter_injuries*6.7*1.31/2+
  loss_data_combined$firefighter_fatalities*6.7*1.31

loss_data_combined<-loss_data_combined %>% 
  mutate(cost = (7705*helicopters + 
                   3635*dozer + 
                   2808*Water_Tenders +
                   5000*engines + 
                   5000*personnel +
                   5000 * crews + 
                   5000*Other_Assigned)/1000000)

write.csv(loss_data_combined, "~/Desktop/californiaUSRA/loss_data_raw.csv")

#select the rows that I will use for analysis and prediction
loss_data_clean<-loss_data_combined %>% 
  select(area = incident_acres_burned,
         longitude = incident_longitude,
         latitude = incident_latitude,
         start_year = start_year,
         start_mon = start_mon,
         duration,
         loss_estimated)
loss_data_clean<-loss_data_clean %>% filter(longitude>=-125,longitude<=-114,
       latitude>=32,latitude<=42.1)
fire_sf <- st_as_sf(loss_data_clean, coords = c("longitude", "latitude"), crs = 4326)
fire_sf<-fire_sf %>% na.omit()

fire_polygon<- fire_sf %>%
  mutate(area_m2 = area * 4046.86,
         radius_m = sqrt(area_m2 / pi))

fire_proj <- st_transform(fire_polygon, crs = 3310)#Project to a CRS with meters

fire_proj <- fire_proj %>%
  mutate(geometry = st_buffer(geometry, dist = radius_m))#Create spatial circles (buffers in meters)

fire_wgs84 <- st_transform(fire_proj, crs = 4326)#Reproject back to EPSG:4326 (lon/lat degrees)
fire_wgs84$duration<-fire_wgs84$duration %>% 
  as.numeric()
write_sf(fire_wgs84, "~/Desktop/californiaUSRA/fire_polygon.geojson")





```




Rasterize the existence and loss data 
```{r}

population_data<-st_read("~/Desktop/californiaUSRA/california population distribution.geojson")

population_data <- st_transform(population_data, st_crs(fire_polygon))
population_vect<-vect(population_data)

template_raster <- rast(ext(CA_vect), resolution = 0.03)

population_raster <- rasterize(population_vect, template_raster,field = "population")
building_raster<-rasterize(population_vect, template_raster,field = "building_amount")
value_raster<-rasterize(population_vect, template_raster,field = "building_value_median")
stack_raster<-c(population_raster,building_raster,value_raster)

fire_polygon<-st_read("~/Desktop/californiaUSRA/fire_polygon.geojson")
fire_polygon$ID<-1:nrow(fire_polygon)

# Initialize list to hold each month's fire/loss rasters
raster_list <- list()

# Unique years and months
year <- unique(fire_polygon$start_year)
mon <- unique(fire_polygon$start_mon)

# Loop through year-month
for (i in year) {
  for (j in mon) {
    
    # Filter fire polygons for this year and month
    temp <- fire_polygon %>% 
      filter(start_year == i, start_mon == j)
    
    if (nrow(temp) > 0) {
      # Convert to SpatVector if not already
      temp_vect <- vect(temp)
      
      # Rasterize fire polygon: 1 = burned, 0 = background
      fire_raster <- rasterize(temp_vect, template_raster, field = 1, background = 0)
      ID_raster <- rasterize(temp_vect, template_raster, field = "ID", background = -999)
      
      # Mask socio rasters to fire area only, exclude the 0 values
      pop_sub      <- mask(population_raster, fire_raster, maskvalues = 0)
      building_sub <- mask(building_raster, fire_raster, maskvalues = 0)
      value_sub    <- mask(value_raster, fire_raster, maskvalues = 0)
      
      # Compute custom weight: population × buildings × value
      cell_weight <- pop_sub * building_sub * value_sub
      
      # Get cell weights as vector
      weight_vals <- values(cell_weight)
      total_weight <- sum(weight_vals, na.rm = TRUE)
      
      # Sum all fire losses in this time period
      loss_val <- sum(temp$loss_estimated, na.rm = TRUE)
      
      # Initialize loss raster with correct shape
      loss_raster <- fire_raster
      
      # Distribute if valid weights exist
      if (!is.na(total_weight) && total_weight > 0) {
        cell_loss_vals <- (weight_vals / total_weight) * loss_val
        values(loss_raster) <- cell_loss_vals
      } else {
        values(loss_raster) <- 0
      }
      
      # Name layers
      name_base <- paste0(i, "-", sprintf("%02d", j))
      names(fire_raster) <- paste0("fire_", name_base)
      names(loss_raster) <- paste0("loss_", name_base)
      
      # Store in list
      raster_list[[paste0(name_base, "_fire")]] <- fire_raster
      raster_list[[paste0(name_base, "_loss")]] <- loss_raster
      raster_list[[paste0(name_base, "_ID")]] <- ID_raster
    }
  }
}

# Combine all rasters into one multi-layer SpatRaster
full_stack <- rast(raster_list)

crs(full_stack)<-crs(CA_vect)
full_stack<-crop(full_stack,CA_vect)
full_stack <- mask(full_stack, CA_vect)

df <- as.data.frame(full_stack, xy = TRUE)

df_long <- pivot_longer(
  df,
  cols = c(ends_with("fire"), ends_with("loss"), ends_with("ID")),
  names_to = "variable",
  values_to = "value"
)

fire_existence <- df %>%
  dplyr::select(x, y, ends_with("_fire"))
fire_existence<-fire_existence %>% 
  pivot_longer(cols = -c(x,y),
               names_to = "time",
               values_to = "value")
fire_existence$year<-substr(fire_existence$time,1,4) %>% 
  as.numeric()
fire_existence$month<-substr(fire_existence$time,6,7) %>% 
  as.numeric()
fire_existence<-fire_existence %>% 
  dplyr::select(-time) %>% 
  filter(year != 1969)

fire_loss<-df %>% 
  dplyr::select(x, y, ends_with("_loss"))
fire_loss<-fire_loss %>% 
  pivot_longer(cols = -c(x,y),
               names_to = "time",
               values_to = "value")
fire_loss$year<-substr(fire_loss$time,1,4)%>% 
  as.numeric()
fire_loss$month<-substr(fire_loss$time,6,7)%>% 
  as.numeric()
fire_loss<-fire_loss %>% 
  dplyr::select(-time)%>% 
  filter(year != 1969)

write_csv(fire_loss, "/Users/nancy/Desktop/californiaUSRA/california_fire_loss.csv")
write_csv(fire_existence, "/Users/nancy/Desktop/californiaUSRA/california_fire_existence.csv")
```






```{r}

temp_prcp_n_elev<-temp_prcp_n_elev %>% 
  mutate(x = round(x,4)) %>% 
  mutate(y = round(y,4))
fire_existence<-fire_existence %>% 
  mutate(x = round(x, 4),
         y = round(y , 4))

fire_existence_data<-fire_existence %>% 
  left_join(temp_prcp_n_elev, by = c("x","y","year","month")) %>% 
  na.omit()

write.csv(fire_existence_data, "/Users/nancy/Desktop/USRA/california fire/california_fire_existence_data.csv")

```


Random forest for fire existence
```{r}
fire_existence_data <- read.csv("/Users/nancy/Desktop/USRA/california fire/california_fire_existence_data.csv")
fire_existence_data$value <- factor(fire_existence_data$value, levels = c(0, 1), labels = c("no", "yes"))
fire_existence_data$log_prcp <- log1p(fire_existence_data$precipitation)  # log(precip + 1)



set.seed(123)
split_index <- createDataPartition(fire_existence_data$value, p = 0.8, list = FALSE)
train_data <- fire_existence_data[split_index, ]
test_data  <- fire_existence_data[-split_index, ]

rf_model <- ranger(value ~ elevation + temperature + log_prcp + month,
                   data = train_data,
                   num.trees = 200,
                   mtry = 2,
                   probability = TRUE, # predicted probabilities
                   min.node.size = 10,
                   class.weights = c("no" = 1, "yes" = 1000))

rf_pred <- predict(rf_model, data = test_data)$predictions
rf_prob_yes <- rf_pred[, "yes"]


threshold <- 0.15
rf_pred_class <- ifelse(rf_prob_yes > threshold, "yes", "no")
rf_pred_class <- as.factor(rf_pred_class)
confusionMatrix(rf_pred_class, test_data$value)

library(pROC)
rf_roc <- roc(response = test_data$value,
              predictor = rf_prob_yes,
              levels = c("no", "yes"))
plot(rf_roc)
text(x = 0.6, y = 0.2, labels = paste0("AUC = ", round(pROC::auc(rf_roc), 5)))
auc(rf_roc)

saveRDS(rf_model, "~/Desktop/californiaUSRA/rf_model_existence.rds")

```

Random forest for loss model
```{r}
fire_loss<-read.csv("/Users/nancy/Desktop/californiaUSRA/california_fire_loss.csv")
natural_factors<-read.csv("/Users/nancy/Desktop/californiaUSRA/california_natural_factors.csv")

fire_loss$value<-fire_loss$value %>% replace_na(0)
fire_existence<-read.csv("/Users/nancy/Desktop/californiaUSRA/california_fire_existence.csv")

population_data<-st_read("~/Desktop/californiaUSRA/california population distribution.geojson")

fire_loss<-fire_loss %>% 
  mutate(x = round(x,4),
         y = round(y,4))
fire_existence<-fire_existence %>% 
  mutate(x = round(x,4),
         y = round(y,4))
natural_factors<-natural_factors%>% 
  mutate(x = round(x,4),
         y = round(y,4))

fire_loss_data<-fire_loss %>% 
  left_join(natural_factors, by = c("x", "y", "year","month")) %>% 
  left_join(fire_existence, by = c("x", "y", "year","month")) %>% 
  na.omit()
fire_loss_data_with_fire <- fire_loss_data %>% 
  filter(value.y == 1)

fire_loss_sf<-st_as_sf(fire_loss_data_with_fire, coords = c("x","y"), crs = 4326)

population_data <- st_transform(population_data, st_crs(fire_loss_sf))
points_with_county <- st_join(fire_loss_sf, population_data, join = st_intersects)

fire_loss_data_completed <- as.data.frame(points_with_county)  
fire_loss_data_completed <- fire_loss_data_completed %>% 
  dplyr::select(temperature, precipitation, elevation, population, building_amount, building_value_median, value.x)
fire_loss_data_completed<-fire_loss_data_completed %>% 
  rename("loss_estimated" = value.x)
fire_loss_data_clean<- fire_loss_data_completed %>% 
  na.omit()

set.seed(123)
split_index <- createDataPartition(fire_loss_data_clean$loss_estimated, p = 0.8, list = FALSE)
train_data <- fire_loss_data_clean[split_index, ]
test_data  <- fire_loss_data_clean[-split_index, ]

rf_model <- ranger(
  loss_estimated ~ ., 
  data = train_data,
  num.trees = 500,
  mtry = 3, # heuristic
  min.node.size = 5,
  importance = "impurity"
)
rf_pred <- predict(rf_model, data = test_data)$predictions

# Metrics
rf_rmse <- rmse(test_data$loss_estimated, rf_pred)
rf_mae  <- mae(test_data$loss_estimated, rf_pred)
rf_r2   <- cor(test_data$loss_estimated, rf_pred)^2

rf_rmse; rf_mae; rf_r2
#11.03405
#1.562829
#0.8726746

saveRDS(rf_model, "~/Desktop/californiaUSRA/rf_model_loss.rds")

```


Prepare the prediction data gained from climate model
```{r}
tmax_rcp85<-rast("~/Desktop/californiaUSRA/tasmax.rcp85.CNRM-CM5.CRCM5-OUR.mon.NAM-22i.raw.nc")
tmin_rcp85<-rast("~/Desktop/californiaUSRA/tasmin.rcp85.CNRM-CM5.CRCM5-OUR.mon.NAM-22i.raw.nc")
prcp_rcp85<-rast("~/Desktop/californiaUSRA/prec.rcp85.CNRM-CM5.CRCM5-OUR.mon.NAM-22i.raw.nc")
elev_rcp85<-rast("~/Desktop/californiaUSRA/orog.rcp85.CNRM-CM5.CRCM5-OUR.fixed.NAM-22i.raw.nc")


CA_vect<-project(CA_vect, crs(tmax_rcp85))
template_raster <- rast(ext(CA_vect), resolution = 0.03)

cropped_raster <- crop(elev_rcp85,CA_vect )
elev_raster <- mask(cropped_raster, CA_vect)
elev_raster <- resample(elev_raster, template_raster, method = "near")
elev_df<-as.data.frame(elev_raster, xy = TRUE)

making_it_dataframe<- function(temp){
  cropped_raster <- crop(temp,CA_vect )
  
  temp_raster <- mask(cropped_raster, CA_vect)
  
  raster_date<-(time(temp_raster))
  
  raster_date<-as.character(raster_date)
  
  temp_raster <- resample(temp_raster, template_raster, method = "near")
  
  temp_df <- as.data.frame(temp_raster, xy = TRUE)
  
  names(temp_df) <- c("x", "y",raster_date)
  
  
  temp_df[,-c(1,2)] <- (temp_df[,-c(1,2)] - 273.15) * 9/5 + 32
  
  temp_df_longer<-temp_df %>% 
    pivot_longer(cols = colnames(temp_df)[-c(1,2)],
                 names_to = "Date",
                 values_to = "Temperature")
  temp_df_longer<-temp_df_longer %>% 
    na.omit()
}
tas_df<-making_it_dataframe(tmax_rcp85)

tmin_df<-making_it_dataframe(tmin_rcp85)

prcp_df<-making_it_dataframe(prcp_rcp85)



prediction_data<-tas_df %>% 
  left_join(tmin_df, by = c("x", "y", "Date")) %>% 
  left_join(prcp_df, by = c("x", "y", "Date")) %>% 
  left_join(elev_df, by = c("x", "y"))

prediction_data<- prediction_data %>% 
  mutate(temperature = (Temperature.x+Temperature.y)/2)

colnames(prediction_data)<-c("x", "y", "Date", "tasmax", "tasmin", "precipitation", "elevation", "temperature")

prediction_data<-prediction_data %>% 
  mutate(year = as.numeric(substr(Date, 1, 4)) ) %>% 
  mutate(month = as.numeric(substr(Date, 6,7))) %>% 
  select(-Date)

prediction_data<-prediction_data %>% 
  filter(year>=2025)
prediction_data$temperature<-(prediction_data$temperature - 32) * 5/9 + 273.15
prediction_data<-prediction_data %>% 
  select(-tasmax,-tasmin)
prediction_data$log_prcp<-log1p(prediction_data$precipitation)

write_csv(prediction_data, "~/Desktop/californiaUSRA/prediction_data_from_climateModel")

```

Predict for future
```{r}
prediction_data<-read.csv("~/Desktop/californiaUSRA/prediction_data_from_climateModel")
population_data<-st_read("~/Desktop/californiaUSRA/california population distribution.geojson")

population_data <- st_transform(population_data, 4326)
rf_model<- readRDS("~/Desktop/californiaUSRA/rf_model_existence.rds")
loss_model<-readRDS("~/Desktop/californiaUSRA/rf_model_loss.rds")

year<-unique(prediction_data$year)
month<-unique(prediction_data$month)
year<-2025:2055
existence_list <- list()
for (i in year) {
  for (j in month) {
    temp <- prediction_data %>%
      filter(year == i, month == j) %>%
      dplyr::select(x, y, month, temperature, precipitation, elevation,log_prcp)
    
    if (nrow(temp) == 0) next
    
    # Predict existence
    rf_pred <- predict(rf_model, 
                       data = temp)$predictions
    
    rf_prob_yes  <- rf_pred[, "yes"]
    rf_pred_class <- ifelse(rf_prob_yes > 0.15, 1, 0)
    
    # Build pred_data with climate vars
    pred_data <- data.frame(
      x = temp$x,
      y = temp$y,
      month = temp$month,
      rf_prob_yes = rf_prob_yes,
      rf_pred_class = rf_pred_class,
      precipitation = temp$precipitation,
      elevation = temp$elevation,
      temperature = temp$temperature
    )
    
    pred_data_sf<-st_as_sf(pred_data, coords = c("x","y"),crs = 4326)
    
    points_with_county <- st_join(pred_data_sf, population_data, join = st_intersects)
    
    coords <- st_coordinates(points_with_county) 
    pred_data_completed<-points_with_county %>% 
      st_drop_geometry()%>% 
      mutate( x = coords[,1],
              y = coords[,2]) %>% 
      na.omit()
    
    pred_data_completed <- pred_data_completed %>% 
      dplyr::select(x, y, month,temperature, precipitation, elevation, population, building_amount, building_value_median, rf_pred_class)
    
    loss_predict <- predict(loss_model, 
                            data = pred_data_completed)$predictions
    
    
    pred_data_completed$loss_predicted<-ifelse(pred_data_completed$rf_pred_class == 1, loss_predict, 0)
    pred_data_completed<-pred_data_completed %>% 
      dplyr::select(x, y , loss_predicted)
    existence_list[[paste0(i,"-",j)]] <- pred_data_completed
    print(paste0(i,"-",j))
  }
}
df<-data.frame()
head(df)

for(i in 1:length(existence_list)){
  temp <- as.data.frame(existence_list[[i]])
  
  if (i == 1) {
    df <- temp
  } else {
    df <- df %>%
      left_join(temp, by = c("x", "y"))
  }
  
}
colnames(df)<-c("x","y",names(existence_list))

df$cummulative_losses <- rowSums(df[, 3:(2 + length(existence_list))], na.rm = TRUE)

ggplot(data = df, aes(x = x, y = y, col = cummulative_losses)) +
  geom_point() +
  scale_color_gradientn(
    colours = c( "yellow", "red"),  # low → mid → high
    name = "Cumulative Losses"
  ) +
  theme_minimal() +
  labs(
    title = "Cumulative Losses from 2025 to 2055",  # main plot title
    x = "Longitude",
    y = "Latitude"
  )

#get the cummulative losses

total_row <- colSums(df[, 3:(2 + length(existence_list))], na.rm = TRUE)

total_row_df <- as.data.frame(total_row)

time_cols <- colnames(df)[3:(2 + length(existence_list))]

time_fixed <- gsub("-(\\d)$", "-0\\1", time_cols) 
time_fixed <- paste0(time_fixed, "-01")            

total_row_df$time <- as.Date(time_fixed, format = "%Y-%m-%d")
total_row_df<-total_row_df %>% 
  mutate(cum_loss = cumsum(total_row))

ggplot(data = total_row_df, aes(x = time, y = cum_loss))+
  geom_line()+
  labs(title = "cummulative losses with respect to time")

ggplot(total_row_df, aes(time, cum_loss)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, linetype = 2, color = "black") +   
  geom_smooth(method = "loess", se = FALSE, color = "red") +                
  labs(title = "cummulative losses vs linear (dashed line) vs losses (red curve)")

```

```{r}
     
```

